---
title: "Lab-03 Replication"
author: "Christopher Prener, Ph.D."
date: '(`r format(Sys.time(), "%B %d, %Y")`)'
output: 
  github_document: default
  html_notebook: default 
always_allow_html: yes
---

## Introduction
This notebook provides a replication of Lab-03.

## Dependencies
This notebook requires the following packages:

```{r load-packages}
# tidyverse packages
library(ggplot2)       # static mapping

# mapping packages
library(mapview)      # preview spatial data
library(sf)           # spatial tools

# other packages
library(here)         # file path management
library(RColorBrewer) # color brewer palettes
library(viridis)      # viridis color palettes
```

We'll also need a custom function that Chris has written for creating map breaks: 

```{r load-functions}
source(here("source", "map_breaks.R"))
```

## Load Data
The data for this assigment are all in the `data/` folder, and include a number of different layers:

```{r load-data}
# city boundary
stl_boundary <- st_read(here("data", "STL_BOUNDARY_City.geojson"))

# water layers
il_hydro <- st_read(here("data", "IL_HYDRO_Mississippi.geojson"))
stl_hydro <- st_read(here("data", "STL_HYDRO_AreaWater.geojson"))

# highways
highways <- st_read(here("data", "STL_TRANS_PrimaryRoads", "STL_TRANS_PrimaryRoads.shp"))

# owner occupied housing
housing <- st_read(here("data", "STL_HOUSING_OwnerOccupied.geojson"))
```

We're now ready to map these data.

## Part 1 - Exploration
In my console, I've used `mapview::mapview()` to explore each of these different data sets. All are polygon data except for the highways, which are line data.

## Part 2 - Mapping
The following code chunk creates map breaks using the "fisher" approach, and then layers data on water features and highways over the census tracts showing the percent of homes that are occupied by their owners. 

```{r, housing-map}
## create breaks
housing <- map_breaks(housing, var = "pct_owner_occupied", newvar = "map_breaks",
                    style = "fisher", classes = 5, dig_lab = 2)

## map binned data
p1 <- ggplot() +
  geom_sf(data = stl_boundary, fill = "#ffffff", color = NA) +
  geom_sf(data = housing, mapping = aes(fill = map_breaks)) +
  geom_sf(data = il_hydro, fill = "#d4f1f9") +
  geom_sf(data = stl_hydro, fill = "#d4f1f9") +
  geom_sf(data = highways, color = "#000000") +
  geom_sf(data = stl_boundary, fill = NA, color = "#2a2a2a", size = .6) +
  scale_fill_brewer(palette = "Greens", name = "% Owner Occupied") +
  labs(
    title = "% of Housing Occupied by Homeowners",
    subtitle = "City of St. Louis, 2010",
    caption = "Data via U.S. Census Bureau\nMap by Christopher Prener, PhD"
  )

p1
```

The map shows that owner occupied rates are highest in the southwest portion of St. Louis. We'll save it using the following code:

```{r, save-plot}
ggsave(plot = p1, filename = here("results", "owner_occupied_housing.png"))
```

The map is now saved for dissemination!
